---
title: "Getting Started with lassoboot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lassoboot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(lassoboot)
```

## What this package does

`lassoboot` answers the question: *which of my predictors actually matter, given
that I measured them with known instrument uncertainty?*

Standard lasso regression assumes predictors are measured exactly. In practice,
every sensor, balance, and test method introduces noise. When that noise is
substantial relative to the predictor range — a 4 % coefficient of variation
on a concrete strength test, a 0.07 wt % standard deviation on an XRF alumina
measurement — it inflates uncertainty in the coefficient estimates and can cause
predictors to be selected or dropped inconsistently across replicates.

`lassoboot` propagates declared measurement uncertainties through a parametric
bootstrap: in each of *B* iterations the predictors are perturbed by their
declared noise, the response is regenerated from the fitted model, and lasso is
refit. The result is a **selection probability** (fraction of iterations a
predictor is retained) and a **stability score** (max selection probability
across the regularization path), both of which are more informative than a
single regularized point estimate.

## A four-predictor example

We will use a small synthetic dataset to illustrate the workflow before
introducing the bundled `concrete` data.

```{r toy-data}
set.seed(101)
n  <- 80
df <- data.frame(
  x1 = rnorm(n),        # strong signal
  x2 = rnorm(n),        # moderate signal
  x3 = rnorm(n),        # noise
  x4 = rnorm(n)         # noise
)
df$y <- 2 * df$x1 + 0.8 * df$x2 + rnorm(n, sd = 0.6)
```

### Step 1: `lb_spec()` — declare the model

```{r spec}
spec <- suppressMessages(
  lb_spec(
    y ~ x1 + x2 + x3 + x4,
    data        = df,
    uncertainty = lb_uncertainty(
      x1 = std(0.10),    # absolute SD of 0.10 units
      x2 = cov(5.0)      # 5 % coefficient of variation
    )
  )
)
spec
```

`lb_spec()` validates the formula, checks that every uncertainty term exists
as a numeric column, infers non-negativity constraints where appropriate, and
stores everything needed for the bootstrap. No fitting happens yet.

### Step 2: `lb_bootstrap()` — run the parametric bootstrap

```{r bootstrap}
boot <- withr::with_seed(42, lb_bootstrap(spec, B = 200))
boot
```

With `B = 200` this takes a few seconds. Production analyses typically use
`B = 1000` or more. The seed is set via `withr::with_seed()` so the global RNG
state is not modified.

### Step 3: `tidy()` — read the results

```{r tidy}
library(dplyr)

td <- tidy(boot)
td
```

Each row is one predictor. The key columns:

| Column | Meaning |
|---|---|
| `estimate` | Bootstrap mean coefficient (zeros included for unselected iterations) |
| `conf.low` / `conf.high` | Bootstrap quantile interval (default 95 %) |
| `selection_prob` | Fraction of *B* iterations in which this term was selected |
| `stability_score` | Max selection probability across the lambda path |

`x1` and `x2` should have high selection probabilities; `x3` and `x4` should
be near zero.

### Step 4: `autoplot()` — visualize

```{r plot-coef}
autoplot(boot)
```

The color encodes selection probability. Terms towards the right with dark
color are robustly selected with positive effects.

```{r plot-selection}
autoplot(boot, type = "selection")
```

The dashed line at 0.5 is the conventional "majority selection" threshold.
Terms above it are selected in more than half of bootstrap iterations.

### Filtering significant terms

```{r filter}
tidy(boot) |>
  filter(lb_is_significant(pick(everything()), method = "selection", threshold = 0.5))
```

`lb_is_significant()` is a convenience predicate for use inside `dplyr::filter()`.
The three available methods — `"ci"` (CI excludes zero), `"selection"`, and
`"stability"` — are discussed in `vignette("interpreting-results")`.

## The `concrete` dataset

The package ships a 75-row dataset of limestone calcined clay cement (LC3)
mortar cube strength measurements:

```{r concrete-glimpse}
dplyr::glimpse(concrete)
```

```{r uncertainty-concrete}
uncertainty_concrete
```

The companion `uncertainty_concrete` tibble declares ASTM-sourced measurement
uncertainties for each continuous predictor. Passing it to `lb_spec()` is
all that is needed to activate uncertainty propagation for those columns.

A full worked example on the `concrete` data appears in
`vignette("interpreting-results")`. The methodology behind measurement
uncertainty propagation and the choice of sigma method is covered in
`vignette("measurement-uncertainty")`.

## Next steps

- `vignette("measurement-uncertainty")` — the three uncertainty types, derived
  quantities, normalization, and how sigma method affects CI coverage.
- `vignette("interpreting-results")` — selection probability vs. CI vs.
  stability score, correlated predictors, prediction grids, and nested
  cross-validation.
- `?lb_control` — all tuning parameters with defaults and rationale.
- `?lb_folds_grouped` — group-aware cross-validation to prevent mixture
  leakage across folds.
