---
title: "Interpreting Bootstrap Lasso Results"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting Bootstrap Lasso Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(lassoboot)
library(dplyr)
```

## Building the concrete dataset model

We will work through a complete example using the bundled `concrete` dataset —
75 mortar cube strength measurements from five calcined clay cement (LC3)
blends at three ages.

```{r spec}
spec <- suppressMessages(
  lb_spec(
    strength_MPa ~ (clayHOH + log(clayHOH)) * replacement * alumina * SSA * D50,
    data        = concrete,
    uncertainty = uncertainty_concrete,
    folds       = lb_folds_grouped("mixture")
  )
)
```

The `folds = lb_folds_grouped("mixture")` argument ensures that all age
measurements for a given mixture are assigned to the same cross-validation
fold — preventing mixture identity from leaking across train/test splits.

```{r bootstrap, cache = FALSE}
boot <- withr::with_seed(2025, lb_bootstrap(spec, B = 300))
```

## Three complementary significance criteria

`tidy()` returns one row per predictor with estimates, confidence intervals,
and three complementary diagnostics:

```{r tidy}
td <- tidy(boot)
select(td, term, estimate, conf.low, conf.high, selection_prob, stability_score) |>
  arrange(desc(selection_prob))
```

### 1. Selection probability

```{r selection-plot}
autoplot(boot, type = "selection", threshold = 0.5)
```

**Selection probability** is the fraction of bootstrap iterations in which a
predictor had a nonzero coefficient at `lambda.min`. It answers: *over repeated
perturbations of the data, how often does lasso choose this predictor?*

A threshold of 0.5 ("majority selection") is a reasonable starting point for
exploratory analysis. Use `lb_is_significant()` to filter:

```{r filter-selection}
td |> filter(lb_is_significant(pick(everything()), method = "selection", threshold = 0.5))
```

### 2. CI exclusion of zero

```{r coef-plot}
autoplot(boot)
```

Bootstrap CIs that exclude zero are a supplementary indicator — a predictor
can have a CI excluding zero but a low selection probability if it is selected
inconsistently (see the correlated-predictor section below). CIs should be
paired with selection probability, not used alone.

```{r filter-ci}
td |> filter(lb_is_significant(pick(everything()), method = "ci"))
```

### 3. Stability score

```{r stability-plot}
autoplot(boot, type = "stability", top_n = 10)
```

**Stability score** is the maximum selection probability across the full
regularization path. A predictor with a high stability score is selected
consistently regardless of the specific lambda value chosen — it is robust to
the arbitrary choice of regularization strength.

Stability is the most conservative criterion. For confirmatory claims in a
publication, require `stability_score > 0.8`.

```{r filter-stability}
td |> filter(lb_is_significant(pick(everything()), method = "stability", threshold = 0.8))
```

### When to use each

| Goal | Recommended criterion |
|---|---|
| Exploratory screening | Selection probability ≥ 0.5 |
| Reducing to a parsimonious model | Stability score ≥ 0.8 |
| Reporting effect sizes | Bootstrap CI, paired with one of the above |
| All three simultaneous | `method = "all"` |

## Gelman 2-SD scaling for comparing effect sizes

Raw lasso coefficients are in the original units of each predictor and cannot
be compared directly when predictors have different scales (SSA in m²/g vs.
replacement in %). The `scale = "gelman"` option in `tidy()` multiplies each
coefficient by `2 * sd(predictor)`, placing all numeric predictors on a
"2-standard-deviation change" scale (Gelman 2008):

```{r gelman}
tidy(boot, scale = "gelman") |>
  select(term, estimate, conf.low, conf.high, selection_prob) |>
  arrange(desc(abs(estimate)))
```

A coefficient of 0.15 on this scale means: changing the predictor from its
mean to one SD above its mean changes the predicted response by about 0.075
units (half of the 2-SD effect). This allows direct comparison of clayHOH
(range ~330–670 J/g) with replacement (range 10–30 %) on the same axis.

```{r gelman-plot}
autoplot(boot, scale = "gelman", filter = "selected")
```

## Correlated predictors

Lasso tends to select one variable from a group of correlated predictors
arbitrarily. `lb_correlated_pairs()` identifies pairs where:

1. Correlation in the original data exceeds `cor_threshold`.
2. Neither predictor alone has a high selection probability.
3. The joint "either is selected" probability is high.

```{r correlated-pairs}
pairs <- lb_correlated_pairs(boot, cor_threshold = 0.7, prob_threshold = 0.5)
pairs
```

If any pairs are flagged, the appropriate action depends on domain knowledge:
- If the two predictors measure the same underlying phenomenon, prefer the one
  with lower measurement uncertainty.
- If they measure distinct aspects, consider including both and noting the
  instability in your write-up.
- The stability score is particularly useful here: if only one of a correlated
  pair has a high stability score, that one is the more robust choice.

## Prediction grids: `lb_grid()` and `lb_plot_prediction()`

`lb_plot_prediction()` wraps `lb_grid()` to produce a ribbon plot of
`.fitted ± CI` vs. a focal predictor, with all other predictors held at their
median:

```{r pred-plot}
lb_plot_prediction(boot, focal = "clayHOH", n = 60)
```

```{r pred-plot-facet}
lb_plot_prediction(boot, focal = "replacement", n = 40) +
  ggplot2::labs(title = "Predicted strength vs. replacement level")
```

The `at = "median"` default (holding non-focal predictors at their observed
median) is appropriate for a marginal effect plot. For per-combination ribbons
(showing one ribbon per unique clay × age combination), call `lb_grid()`
directly with `at = "observed"` and build the ggplot manually:

```{r grid-observed, eval = FALSE}
grid <- lb_grid(boot, focal = "clayHOH", n = 60, at = "observed")
# grid has one predicted row per (clayHOH value × observed combination)
```

**Extrapolation warning.** `lb_grid()` clips grid values to the observed range
by default. Pass `extrapolate = TRUE` to extend beyond the data; mark any
extrapolated region clearly in the final plot.

## Nested cross-validation: `lb_folds_nested()`

When the data has a nested structure — multiple replicates per mixture, each
mixture at multiple ages — a plain k-fold CV can leak mixture identity into the
test fold. Use `lb_folds_nested()` to prevent this:

```{r nested-folds, eval = FALSE}
spec_nested <- suppressMessages(
  lb_spec(
    strength_MPa ~ (clayHOH + log(clayHOH)) * replacement * alumina,
    data  = concrete,
    uncertainty = uncertainty_concrete,
    folds = lb_folds_nested(
      outer = "mixture",   # whole mixtures in same fold
      inner = "age",       # balance age distribution within each fold
      k_outer = 5
    )
  )
)
```

`lb_folds_grouped("mixture")` (used above) is the simple version: all rows for
a given mixture are assigned to the same fold. `lb_folds_nested()` additionally
stratifies *within* each fold by the `inner` variable so that each fold sees a
balanced age distribution.

For the `concrete` dataset with only 25 unique mixtures and 5-fold CV, the
difference between grouped and nested is small. The effect is largest in
datasets where some combinations are rarer (e.g. missing ages for certain
mixtures).

## Model complexity

```{r complexity}
autoplot(boot, type = "complexity")
```

The complexity histogram shows the distribution of how many predictors were
selected across bootstrap iterations. A wide distribution (spanning, say,
2–12 predictors) suggests the model is on the border of the bias-variance
trade-off and the lambda selection is sensitive to the data. A tight
distribution centered on a small number is reassuring.

## References

Gelman, A. (2008). Scaling regression inputs by dividing by two standard
deviations. *Statistics in Medicine*, **27**(15), 2865–2873.

Meinshausen, N. and Bühlmann, P. (2010). Stability selection.
*Journal of the Royal Statistical Society: Series B*, **72**(4), 417–473.
