---
title: "Measurement Uncertainty in Parametric Bootstrap Lasso"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Measurement Uncertainty in Parametric Bootstrap Lasso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(lassoboot)
library(dplyr)
```

> **Upgrading from v0.1?** Column names in `tidy()` output changed in v0.2.0:
> `estimate` → `mean`, `conf.low`/`conf.high` → `q025`/`q975`,
> `std.error` → `sd`. The `conf.level` argument is now `probs = c(0.025, 0.975)`.
> Use `lb_filter_stable()` instead of `lb_is_significant()`.
> See `vignette("migration-to-v020")` for the full API rename table, and
> `vignette("methodological-foundations")` for the v0.2.0 conceptual framing.

## Why measurement uncertainty matters in lasso

Lasso regression assumes predictors are observed without error. When predictors
carry instrument noise, two problems arise:

1. **Bias toward zero.** Noise in predictors attenuates coefficient estimates
   toward zero, making truly important predictors look weaker than they are.
2. **Unstable selection.** A predictor near the lasso penalty boundary may be
   selected or dropped depending on which side of the noise it falls on a given
   day.

The parametric bootstrap in `lassoboot` addresses both by explicitly modelling
the noise. In each iteration, each predictor is perturbed by an independent
draw from its declared uncertainty distribution before lasso is refit. Over
*B* iterations, the selection probability and coefficient distribution reflect
both model uncertainty *and* measurement uncertainty.

## Declaring uncertainties: the three types

Uncertainty is declared as a tibble via `lb_uncertainty()`. Three types are
supported:

| Type | Argument form | Interpretation |
|---|---|---|
| `"std"` | `std(value)` | Absolute standard deviation in the units of the column |
| `"cov"` | `cov(value)` | Coefficient of variation as a percentage (e.g. `cov(4.0)` = 4 %) |
| `"rel"` | `rel(value)` | Relative SD as a fraction (e.g. `rel(0.04)` = 4 %) |

### Example: ASTM-documented uncertainties

The `uncertainty_concrete` dataset bundled with the package declares
uncertainties for the `concrete` dataset's predictors, sourced from ASTM
standard precision statements:

```{r uncertainty-table}
uncertainty_concrete
```

**Choosing the right type:**
- Use `"std"` when the standard specifies an absolute repeatability value
  (e.g. ASTM C114 XRF alumina: ±0.07 wt %).
- Use `"cov"` when the standard specifies a percent repeatability
  (e.g. ASTM C109 strength: 3 % single-operator CoV).
- Use `"rel"` when you have a fractional relative SD (e.g. from a
  certificate of analysis that reports `u_rel = 0.04`).

A missing column (column exists in data but is not listed in the uncertainty
spec) is treated as having zero uncertainty — the column is not perturbed.
Declaring a non-existent column is a hard error; silently dropping uncertainty
would invalidate the analysis.

## Common ASTM measurement uncertainties

For convenience, typical values from standards relevant to cementitious
materials:

| Measurement | Method | Uncertainty | Type |
|---|---|---|---|
| Al₂O₃ by XRF | ASTM C114 | ±0.07 wt % | `std` |
| SiO₂ by XRF | ASTM C114 | ±0.13 wt % | `std` |
| Compressive strength | ASTM C109 | 3.0 % | `cov` |
| Heat of hydration | ASTM C1897 | ±3.2 J/g | `std` |
| Specific surface area (BET) | ISO 9277 | 2–5 % | `cov` |
| Particle size D50 (laser) | ISO 13320 | 3–5 % | `cov` |
| Loss on ignition | ASTM C114 | ±0.12 wt % | `std` |
| Water demand (Blaine fineness) | ASTM C204 | ±8 m²/kg | `std` |

Always prefer the *within-laboratory repeatability* (single-operator) value
over the *reproducibility* (multi-lab) value when your analysis uses data from
a single lab.

## Derived quantities: `lb_derive()`

When a predictor is computed from other measured quantities (e.g. volume from
mass and density), the derivation must happen *after* uncertainty injection.
Applying uncertainty to the inputs and then recomputing the derived quantity
correctly propagates the correlated noise through the calculation.

```{r derive-example, eval = FALSE}
spec <- lb_spec(
  strength_MPa ~ volume_Alite + volume_Belite + alumina,
  data        = concrete_xrd,    # hypothetical XRD dataset
  uncertainty = lb_uncertainty(
    mass_Alite   = std(0.74, "ASTM C1365 Table 1, s_within"),
    density_Alite = std(0.012, "ASTM C188"),
    mass_Belite   = std(0.68, "ASTM C1365 Table 1, s_within"),
    density_Belite = std(0.010, "ASTM C188")
  ),
  derive = list(
    lb_derive(volume_Alite  = mass_Alite  / density_Alite),
    lb_derive(volume_Belite = mass_Belite / density_Belite)
  )
)
```

The bootstrap loop applies uncertainty to `mass_Alite`, `density_Alite`, etc.,
then recomputes `volume_Alite` from the perturbed inputs. The derive list runs
in declaration order, so chained derives (e.g. `volume_total = volume_Alite +
volume_Belite`) work as expected.

**Constraint interaction.** Default constraints clip numeric columns to
`[0, Inf)` if all observed values are non-negative. Clipping is applied to the
raw measured columns *before* derived quantities are computed. A derived ratio
(like volume = mass/density) is not re-clipped after derivation.

## Normalizing a ratio response: `lb_normalize()`

When the response is the ratio of a test measurement to a reference
(e.g. strength relative to a control mixture), both the test measurement and
the reference carry uncertainty. `lb_normalize()` registers a derived column
and rewrites the formula's left-hand side:

```{r normalize-example, eval = FALSE}
spec <- lb_spec(
  strength_MPa ~ (clayHOH + log(clayHOH)) * replacement * alumina * SSA * D50,
  data        = concrete_with_ref,   # dataset with meas_strength + ref_strength
  uncertainty = lb_uncertainty(
    meas_strength = cov(3.0, "ASTM C109"),
    ref_strength  = cov(2.2, "ASTM C109, mean of 3")
  )
) |>
  lb_normalize(
    response  = strength_MPa,
    measured  = meas_strength,
    reference = ref_strength
  )
```

Each bootstrap iteration perturbs `meas_strength` and `ref_strength`
independently, then recomputes the ratio. The user is responsible for declaring
the uncertainty appropriate to how the values were pre-aggregated: if
`ref_strength` is the mean of three replicates, the declared uncertainty should
be the standard error of that mean (single-measurement SD / sqrt(3)), not the
single-measurement SD.

## Constraints: `lb_constraints()`

By default, `lb_spec()` infers a non-negativity constraint for any predictor
where all observed values are ≥ 0, and no constraint for predictors that cross
zero. A one-time message reports the inferred constraints.

Override defaults explicitly:

```{r constraints-example, eval = FALSE}
spec <- lb_spec(
  y ~ x1 + x2 + x3 + moisture,
  data = df,
  constraints = lb_constraints(
    x1       = c(0, 100),    # bounded weight percent
    x2       = c(0, Inf),    # non-negative, no upper bound
    moisture = c(-Inf, Inf)  # signed; no clipping
  )
)
```

Explicit declaration overrides inference. Columns not listed use the inferred
default. A column declared with `c(-Inf, Inf)` is never clipped even if all
observed values are non-negative.

## The residual sigma method: why the default matters

The parametric bootstrap generates the response in each iteration as:

```
y_star = fitted_original + rnorm(n, 0, sigma_hat)
```

How `sigma_hat` is estimated is controlled by `lb_control(sigma_method = ...)`:

| Method | Description | Prediction-band width |
|---|---|---|
| `"refit"` (default) | OLS on lasso-selected variables; residual SE from unregularized fit | Closest to nominal |
| `"naive"` | `sd(y - fitted_lasso)` | Under-estimates (residuals are shrunk) |
| `"cv"` | Out-of-fold CV residuals | Honest but slow |

The `"refit"` default is chosen because lasso shrinks its own residuals — using
the regularized residual SE underestimates the true noise and results in
bootstrap CIs that are too narrow. Refitting OLS on the selected variables
largely closes this gap at the cost of one additional `lm()` call per fit.

### Coverage simulation

The code below verifies this on simulated data (skipped in package build;
set `LASSOBOOT_COV_VIGNETTE=1` to run):

```{r coverage-sim, eval = nchar(Sys.getenv("LASSOBOOT_COV_VIGNETTE")) > 0}
set.seed(2025)
n_sim   <- 100
n       <- 120
p_total <- 20
p_signal <- 5
B        <- 150

cover_refit <- cover_naive <- numeric(n_sim)

for (i in seq_len(n_sim)) {
  # Simulated sparse data
  X     <- matrix(rnorm(n * p_total), n, p_total)
  beta  <- c(rep(2, p_signal), rep(0, p_total - p_signal))
  y     <- X %*% beta + rnorm(n)
  df_i  <- as.data.frame(X)
  names(df_i) <- paste0("x", seq_len(p_total))
  df_i$y <- as.numeric(y)

  spec_r <- suppressMessages(lb_spec(y ~ ., data = df_i,
    control = lb_control(sigma_method = "refit", cv_reps = 2L,
                         n_lambda = 20L, store_path = FALSE)))
  spec_n <- suppressMessages(lb_spec(y ~ ., data = df_i,
    control = lb_control(sigma_method = "naive", cv_reps = 2L,
                         n_lambda = 20L, store_path = FALSE)))

  boot_r <- withr::with_seed(i, lb_bootstrap(spec_r, B = B))
  boot_n <- withr::with_seed(i, lb_bootstrap(spec_n, B = B))

  td_r   <- tidy(boot_r)
  td_n   <- tidy(boot_n)

  # Check CI coverage for x1 (true coef = 2)
  cover_refit[i] <- td_r$conf.low[td_r$term == "x1"] <= 2 &
                    td_r$conf.high[td_r$term == "x1"] >= 2
  cover_naive[i] <- td_n$conf.low[td_n$term == "x1"] <= 2 &
                    td_n$conf.high[td_n$term == "x1"] >= 2
}

cat("sigma='refit'  empirical 95% CI coverage:", mean(cover_refit), "\n")
cat("sigma='naive'  empirical 95% CI coverage:", mean(cover_naive), "\n")
```

In repeated experiments the `"refit"` method consistently achieves closer to
nominal 95 % coverage than `"naive"`. The gap is most pronounced when the
selected model is sparse relative to the true signal (lasso aggressively
shrinks residuals).

## A note on selective inference

The bootstrap CIs produced by `lassoboot` are **frequency-bootstrap** intervals,
not selectively-inferentially valid intervals in the sense of Lockhart et al.
(2014) or Taylor & Tibshirani (2015). This means:

- The nominal 95 % CI reflects the bootstrap sampling distribution, not a
  post-selection corrected pivot.
- For predictors selected in almost every iteration (`selection_prob ≈ 1`), the
  CI is reliable as a rough effect-size guide.
- For predictors near the selection boundary (`selection_prob ≈ 0.5`), the CI
  is less reliable; treat the selection probability and stability score as the
  primary inferential claim.

The package is designed to lead with selection/stability and treat CIs as
supplementary, which is the appropriate posture for exploratory variable
selection in materials science. Do not report bootstrap CIs as if they were
p-value-corrected confidence intervals for post-selection inference.

## References

Lockhart, R., Taylor, J., Tibshirani, R. J., and Tibshirani, R. (2014).
A significance test for the lasso. *Annals of Statistics*, **42**(2), 413–468.

Meinshausen, N. and Bühlmann, P. (2010). Stability selection.
*Journal of the Royal Statistical Society: Series B*, **72**(4), 417–473.
