---
title: "Methodological Foundations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methodological Foundations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lassoboot)
```

## The problem this package solves

A researcher collects 75 mortar cube strength measurements from five calcined
clay cement blends at three curing ages. She has four candidate predictors:
clay heat of hydration (measured by calorimetry), alumina content (by XRF),
specific surface area (by BET nitrogen adsorption), and median particle
diameter (by laser diffraction). She wants to know which predictors matter.

This is not a large-n problem for which standard lasso asymptotics apply. It
is a **lonely-island problem**: a single dataset, measured with instruments
that have documented precision limits, analysed once. There is no population
of LC3 studies to average over. The uncertainty in "which predictors matter"
comes from two distinct sources:

1. **Model uncertainty** — finite sample size means the estimated lasso path
   is noisy. A different 75-cube experiment would give a different selected
   set.
2. **Measurement uncertainty** — every sensor adds noise. If alumina content
   were measured on a different calibration day, the perturbed values would
   shift the lasso penalty boundary for that variable.

Both sources are real. Ignoring measurement uncertainty understates how
fragile the selected model is. `lassoboot` propagates both simultaneously.

---

## The three methodological pillars

### 1. GUM Monte Carlo for measurement uncertainty propagation

The *Guide to the Expression of Uncertainty in Measurement* (JCGM 100:2008;
"GUM") specifies Monte Carlo simulation as a Supplement-1 method for
propagating uncertainties through non-linear functions. The GUM approach:

- Each measurand carries a declared probability distribution reflecting its
  measurement uncertainty.
- In each trial, all measurands are independently drawn from their
  distributions and the quantity of interest is computed.
- The distribution of results over many trials is the propagated uncertainty.

`lassoboot` implements this exactly: in each of *B* bootstrap iterations,
every predictor declared in `lb_uncertainty()` is perturbed by an independent
draw from its declared distribution (`"std"` → N(0, σ²), `"cov"` → N(0,
(|x|·cv/100)²), `"rel"` → N(0, (|x|·f)²)). The perturbed data are clipped
to declared constraints, derived quantities are recomputed, and lasso is
refit. The *B* refits collectively represent the propagated uncertainty.

This is a **forward propagation**: we do not attempt to back-compute which
component of uncertainty matters most. We simulate the complete analysis as it
would run on slightly different measurements, because that is what "measurement
uncertainty" means in practice.

### 2. SIMEX and bias under measurement error

In ordinary lasso, predictors measured with noise cause **attenuation bias**:
the coefficient estimate for a noisy predictor is biased toward zero because
part of its variance is noise that the model cannot use. Lasso compounds this
by also shrinking coefficients explicitly via the penalty.

SIMEX (simulation extrapolation; Cook and Stefanski 1994) estimates this bias
by fitting the model at progressively amplified noise levels and extrapolating
back to zero noise. The `lassoboot` bootstrap does not implement SIMEX
extrapolation explicitly; instead, it propagates the *declared* noise level
and reports how the selected model varies across draws. This is a more
conservative answer: it asks "how stable is the *analysis* we actually ran,
given the noise we actually had?" rather than "what would the result look like
without noise?"

The practical implication is that `lassoboot` selection probabilities reflect
both model uncertainty and measurement uncertainty simultaneously. A predictor
near the penalty boundary may have a low selection probability not because it
is uninformative but because its SNR (signal-to-noise ratio given its declared
uncertainty) is too low to push it reliably above the lasso penalty. That is
useful diagnostic information: it says "if you could measure this predictor
more precisely, selection stability would improve."

### 3. Meinshausen–Bühlmann stability selection

Meinshausen and Bühlmann (2010) showed that selection stability — the
probability that a predictor is selected across random subsamples — is a more
reliable criterion for variable selection than a single lasso path. Their
analysis produced finite-sample bounds on the expected number of false
discoveries when using a stability threshold.

`lassoboot` generalises this idea: instead of subsampling the observations, it
perturbs the predictors by their declared measurement distributions. This makes
the bootstrap more informative for the lonely-island case, where the sample is
fixed but the measurements are noisy.

Two stability diagnostics are reported by `tidy()`:

- **`selection_prob`**: the fraction of *B* iterations in which the predictor
  had a nonzero coefficient at the CV-optimal lambda. This is the primary
  selection criterion for exploratory variable selection.
- **`stability_score`**: the *maximum* selection probability across the full
  regularization path (cf. Meinshausen and Bühlmann's Π̂). A predictor with
  a high stability score is selected consistently regardless of which lambda
  value is used — it is robust to the arbitrary choice of regularization
  strength. Use stability score as a more conservative criterion for
  confirmatory claims.

Neither metric is a p-value. They are **stability diagnostics**: empirical
summaries of how consistently the lasso identifies a given predictor as
informative under realistic perturbations of the data. A selection probability
of 0.90 means the predictor is selected in 90 % of bootstrap iterations;
it does not mean there is a 10 % probability the predictor is truly zero.

---

## Why not Bayesian lasso?

The Bayesian lasso (Park and Casella 2008; Hans 2009) places a Laplace (double-
exponential) prior on the coefficients and uses the posterior as the
uncertainty representation. This is an elegant formulation but it is not
what `lassoboot` is for, for three reasons:

1. **Measurement uncertainty is not parameter uncertainty.** The Bayesian
   posterior quantifies uncertainty about the true regression coefficients
   given the observed data. Measurement uncertainty is uncertainty about the
   *data themselves* — the predictors were not measured exactly. These are
   different modelling layers. Folding instrument precision into a prior
   on coefficients confounds them.

2. **Prior elicitation is hard in small-n settings.** With *n* = 75 and
   *p* = 20+ interaction terms, the posterior is sensitive to the prior.
   The Laplace prior hyperparameter controls shrinkage globally; `lassoboot`'s
   measurement uncertainty declarations are per-column, sourced from
   normative standards, and do not require subjective tuning.

3. **Selection is not posterior probability.** The Bayesian lasso does not
   produce exactly-zero coefficients (it uses a continuous prior). Variable
   selection requires an additional spike-and-slab layer (Ishwaran and Rao
   2005) or a continuous horseshoe-type prior (Carvalho et al. 2010), each of
   which requires its own posterior computation and threshold choice. The
   `lassoboot` selection probability is a simple, interpretable empirical
   frequency.

The parametric bootstrap occupies a useful middle ground: it uses
frequentist lasso for coefficient estimation (with well-understood regularized
path algorithms), adds genuine uncertainty propagation for the measurement
component, and reports stability diagnostics that are no harder to interpret
than a proportion.

---

## Why this isn't overcomplicated

The full `lassoboot` workflow is:

```r
spec <- lb_spec(y ~ x1 + x2 + x3,
                data        = df,
                uncertainty = lb_uncertainty(x1 = std(0.1), x2 = cov(5.0)))
boot <- lb_bootstrap(spec, B = 500)
tidy(boot)
autoplot(boot)
```

The complexity in this vignette describes *why* those four lines work, not
what the user needs to know to run them. The package has three levels of
user interaction:

1. **Standard use**: declare `lb_spec()` + `lb_bootstrap()`, interpret
   `tidy()` and `autoplot()`. No knowledge of GUM or stability selection
   required.
2. **Tuning**: adjust `lb_control()` parameters (CV folds, sigma method,
   lambda strategy). The defaults were chosen to work well on small-n
   materials-science datasets.
3. **Advanced**: use `lb_derive()` and `lb_normalize()` for computed
   predictors and ratio responses; use `lb_grid()` / `lb_plot_envelopes()`
   for custom prediction grids.

The methodological foundations described here justify the default choices and
allow expert users to decide when non-defaults are appropriate. They are not
prerequisites for using the package.

---

## What `lassoboot` does NOT do

- It does not provide post-selection inference in the sense of Lockhart et al.
  (2014). The bootstrap CIs are frequency-bootstrap intervals, not
  selectively-corrected pivots.
- It does not model measurement error via a latent-variable structure (errors-
  in-variables / EIV regression). It propagates declared distributions forward.
- It does not handle non-Gaussian responses. Gaussian linear lasso only.
- It does not support the elastic net penalty (`alpha < 1`). Lasso only.

See `vignette("migration-to-v020")` for the v0.1 → v0.2.0 API migration
table and `vignette("measurement-uncertainty")` for the uncertainty declaration
syntax.

---

## References

Carvalho, C. M., Polson, N. G., and Scott, J. G. (2010). The horseshoe
estimator for sparse signals. *Biometrika*, **97**(2), 465–480.

Cook, J. R. and Stefanski, L. A. (1994). Simulation extrapolation estimation
in parametric measurement error models. *Journal of the American Statistical
Association*, **89**(428), 1314–1328.

Hans, C. (2009). Bayesian lasso regression. *Biometrika*, **96**(4), 835–845.

Ishwaran, H. and Rao, J. S. (2005). Spike and slab variable selection:
frequentist and Bayesian strategies. *Annals of Statistics*, **33**(2),
730–773.

Joint Committee for Guides in Metrology (2008). *Evaluation of Measurement
Data — Guide to the Expression of Uncertainty in Measurement (GUM).*
JCGM 100:2008. BIPM, Sèvres.

Lockhart, R., Taylor, J., Tibshirani, R. J., and Tibshirani, R. (2014).
A significance test for the lasso. *Annals of Statistics*, **42**(2), 413–468.

Meinshausen, N. and Bühlmann, P. (2010). Stability selection.
*Journal of the Royal Statistical Society: Series B*, **72**(4), 417–473.

Park, T. and Casella, G. (2008). The Bayesian lasso.
*Journal of the American Statistical Association*, **103**(482), 681–686.
