---
title: "Migrating from lassoboot v0.1 to v0.2.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from lassoboot v0.1 to v0.2.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(lassoboot)
library(dplyr)
```

## What lassoboot actually computes

lassoboot runs a **parametric bootstrap** over a lasso regression. In each of
*B* iterations:

1. Each declared predictor is perturbed by an independent draw from its
   measurement uncertainty distribution.
2. The response is regenerated as `ŷ + ε`, where `ŷ` is the fitted value from
   the *original* lasso fit and `ε ~ N(0, σ̂²)`.
3. Lasso is refit on the perturbed data.

The *B* refits produce a distribution of coefficient vectors. Everything
`lassoboot` reports — selection probabilities, stability scores, quantile
intervals, prediction envelopes — is a summary of that distribution.

### v0.2.0 conceptual reframe

In v0.1, `lassoboot` was framed primarily as a tool for **coefficient
inference**: which predictors have non-zero effects, with what uncertainty?
The tidy output mimicked `broom::tidy()` conventions (`estimate`, `conf.low`,
`conf.high`, `std.error`) to invite comparison with OLS t-based intervals.

In v0.2.0, the primary framing shifts to **prediction envelopes**: given a
new combination of predictor values, where will the response fall, and how
much of that uncertainty is epistemic (we don't know the model well) vs.
aleatoric (irreducible scatter)? Selection probabilities are now explicitly
described as **stability diagnostics** — diagnostic evidence that a predictor
is robustly identified — not as a substitute for p-values.

This reframe does not change any computation. It changes how the outputs
are named, documented, and plotted.

---

## Column name changes in `tidy()`

The `tidy()` method for `lb_boot` objects returns one row per predictor. In
v0.2.0, column names are:

| v0.1 name | v0.2.0 name | Notes |
|---|---|---|
| `estimate` | `mean` | Bootstrap mean (zeros from unselected iterations included) |
| `estimate_median` | `median` | Bootstrap median |
| `std.error` | `sd` | Bootstrap standard deviation |
| `conf.low` | `q025` | 2.5 % quantile (by default; controlled by `probs`) |
| `conf.high` | `q975` | 97.5 % quantile |
| `selection_prob` | `selection_prob` | Unchanged |
| `n_selected` | `n_selected` | Unchanged |
| `stability_score` | `stability_score` | Unchanged |

The `conf.level` argument is replaced by `probs`, a numeric vector of
quantile probabilities. The default `probs = c(0.025, 0.975)` gives the same
interval as the old `conf.level = 0.95`.

**There is no backward-compatibility aliasing.** Code that references
`td$conf.low` will error immediately with a clear message from dplyr. This is
intentional: silent wrong-column access is worse than an early error.

### Updating existing code

```r
# v0.1
td <- tidy(boot, conf.level = 0.90)
td$estimate
td$conf.low
td$conf.high

# v0.2.0
td <- tidy(boot, probs = c(0.05, 0.95))
td$mean
td$q050   # quantile name is derived from probs × 1000, zero-padded to 3 digits
td$q950
```

The quantile column names are always three-digit zero-padded integers:
`q025`, `q975`, `q050`, `q950`, `q010`, `q990`, etc.

---

## Filtering: `lb_filter_stable()` replaces `lb_is_significant()`

`lb_is_significant()` is retained as a **deprecated** wrapper and will be
removed in a future version. Migrate to `lb_filter_stable()`:

```r
# v0.1
tidy(boot) |>
  filter(lb_is_significant(pick(everything()), method = "selection", threshold = 0.5))

# v0.2.0
lb_filter_stable(tidy(boot), min_selection_prob = 0.5)
```

`lb_filter_stable()` accepts three independent criteria (any combination):

| Argument | Meaning |
|---|---|
| `min_selection_prob` | Keep rows where `selection_prob ≥` this value |
| `min_stability_score` | Keep rows where `stability_score ≥` this value |
| `quantiles_exclude_zero` | Keep rows where `q025 > 0` or `q975 < 0` |

All supplied criteria are AND-ed together. Rows must satisfy every declared
condition to be retained.

---

## Prediction intervals: `interval` argument

`predict.lb_boot()`, `lb_grid()`, `lb_plot_prediction()`, and
`lb_plot_envelopes()` now support an `interval` argument:

**`interval = "confidence"` (default):** The bootstrap quantile envelope
around the *mean response*. Captures uncertainty in the regression surface —
how precisely the model is estimated. Narrows as *n* grows.

**`interval = "prediction"`:** Adds residual scatter `ε ~ N(0, σ̂²)` to each
bootstrap draw before computing quantiles. Captures where a *new individual
observation* will fall. Does not narrow with *n* beyond the point where σ̂
dominates.

**`interval = "both"`** (in `lb_plot_envelopes()` only): draws both ribbons
simultaneously — prediction ribbon in the background at lower opacity,
confidence ribbon in the foreground at higher opacity.

```{r interval-example, eval = FALSE}
boot <- lb_bootstrap(lb_spec(y ~ x1 + x2, data = df), B = 300)

# Confidence band
lb_plot_prediction(boot, focal = "x1", interval = "confidence")

# Prediction band
lb_plot_prediction(boot, focal = "x1", interval = "prediction")

# Both (envelope plot)
lb_plot_envelopes(boot, focal = "x1", interval = "both")
```

---

## Measurement uncertainty precision levels

`lb_uncertainty()` now supports two precision levels per term via the `multi`
argument:

```{r multi-precision}
u <- lb_uncertainty(
  alumina      = std(0.071, source = "ASTM C114"),          # single-lab
  strength_MPa = cov(3.0,  multi = 6.0, source = "ASTM C109")  # single + multi-lab
)
u
```

The `lb_control(precision = ...)` argument selects which column is used during
the bootstrap:

- `precision = "single"` (default): use `value_single` — within-laboratory
  repeatability. Appropriate when all data come from one lab.
- `precision = "multi"`: use `value_multi` — between-laboratory
  reproducibility. Use when comparing results across labs or for worst-case
  sensitivity analysis.

**Backward compatibility.** The old `uncertainty_concrete` dataset uses a
single `value` column (the v0.1 format). `lb_uncertainty()` automatically
backfills `value_single = value_multi = value` when it reads a legacy tibble,
so existing code continues to work without modification.

---

## `lb_normalize()` is now exported

`lb_normalize()` was an internal function in v0.1. It is now exported and
documented. Use it when the response is a ratio of two measured quantities:

```{r normalize-example, eval = FALSE}
# lb_spec() requires the response column to exist in data at build time.
# Initialise a placeholder; lb_normalize() overwrites it with the ratio.
df_with_ref$strength_ratio <- NA_real_

spec <- lb_spec(
  strength_ratio ~ clayHOH * replacement,
  data        = df_with_ref,
  uncertainty = lb_uncertainty(
    meas_strength = cov(3.0, "ASTM C109"),
    ref_strength  = cov(2.2, "ASTM C109")
  )
) |>
  lb_normalize(
    response  = strength_ratio,
    measured  = meas_strength,
    reference = ref_strength
  )
```

Each bootstrap iteration perturbs `meas_strength` and `ref_strength`
independently before recomputing the ratio, correctly propagating their
correlated uncertainties.

---

## Summary of API changes

| Feature | v0.1 | v0.2.0 |
|---|---|---|
| tidy column: mean | `estimate` | `mean` |
| tidy column: lower CI | `conf.low` | `q025` (or `q` + probs × 1000) |
| tidy column: upper CI | `conf.high` | `q975` |
| tidy column: SD | `std.error` | `sd` |
| tidy interval control | `conf.level = 0.95` | `probs = c(0.025, 0.975)` |
| filter helper | `lb_is_significant()` | `lb_filter_stable()` |
| predict interval | not supported | `interval = "confidence"/"prediction"` |
| two precision levels | not supported | `lb_uncertainty(..., multi = ...)` |
| precision selection | not supported | `lb_control(precision = "single"/"multi")` |
| `lb_normalize()` | internal | exported |
| `lb_plot_envelopes()` | not available | new in v0.2.0 |

---

## References

Efron, B. and Tibshirani, R. J. (1993). *An Introduction to the Bootstrap.*
Chapman & Hall.

Gelman, A. (2008). Scaling regression inputs by dividing by two standard
deviations. *Statistics in Medicine*, **27**(15), 2865–2873.

Meinshausen, N. and Bühlmann, P. (2010). Stability selection.
*Journal of the Royal Statistical Society: Series B*, **72**(4), 417–473.
